\documentclass{asl}
%
\title{Review: Geoffrey Hellman, \emph{Mathematics and its Logics}}
%
\author{Chris Scambler}
\revauthor{C. Scambler}
\address{All Souls College\\
University of Oxford}
\email{cscambler@gmail.com}
%
\usepackage{tikz}
\usepackage{bussproofs}
% These will be typeset in italics
\newtheorem{Theorem}{Theorem}[section]
\newtheorem{Proposition}[Theorem]{Proposition}
\newtheorem{Lemma}[Theorem]{Lemma}
\newtheorem{Corollary}[Theorem]{Corollary}
\usepackage{times}
% These will be typeset in Roman
\theoremstyle{definition}
\newtheorem{Definition}[Theorem]{Definition}
\newtheorem{Fact}[Theorem]{Fact}
\newtheorem{Conjecture}[Theorem]{Conjecture}
\newtheorem{Remark}[Theorem]{Remark}
%
\begin{document} 
\maketitle
%\begin{abstract} \end{abstract}

\section{Introduction}

\emph{Mathematics and its logics} is a collection of Geoffrey Hellman's papers. 
The volume covers a huge variety of topics:
there are essays on the foundational role of category theory, 
on the nature and value of restrictive programs in foundations of mathematics such as nominalism and predicativism, 
on modal structuralism, on the Quine-Carnap debate, 
on intuitionism and smooth infinitesimal analysis, 
and much more besides. 
Hellman's important contributions on all of these topics are reflected in the collected papers, 
and in addition they tend to contain excellent introductions to the relevant debates along the way.
As a result, 
the book would make a valuable addition to the library of any graduate student or professional with a specialization in philosophy of mathematics. 

The book is split into three parts reflecting different aspects of Hellman's work. 
Each part consists of a series of papers with a common theme.
Most of the papers are very well known, have been highly influential, 
and have already been subjects of much discussion in the literature.
However, in addition, 
each part also offers a newly written paper.
One of these is a critical piece, responding to recent work by Penelope Maddy. 
The other two push at the frontiers of the existing work showcased in parts one and two.
Since the older works are already so well known and have been so extensively discussed, 
after summarizing the papers offered in each part, 
in this review I will go on to discuss the contrubtions made by the two new non-critical papers.

\section{Overview}

Part I consists in papers related to Hellman's modal structuralism and surrounding issues.
The first paper of part I, \emph{Structuralism without Structures}, 
is one of the founding works in the theory of modal structuralism;
in it, mathematical foundations are developed in detail for many areas of mathematics
in modal structuralist terms, and the philosophical credentials of modal structuralism are assessed.
(The paper also contains interesting points of comparison between modal structuralism and the predicativst viewpoint, 
which is itself the topic of scrutiny in part II.) 
The second paper, 
\emph{What Is Categorical Structuralism?}, 
presents a comprehensive discussion of the role of category theory in foundations of mathematics, 
arguing 
(against Colin McClarty and Steve Awodey) 
that ultimately category theory is in an important sense lacking in this regard. 
Papers three and four, 
\emph{On the Significance of the Burali Forti Paradox} 
and \emph{Extending the Iterative Concept of Set}, 
discuss the set-theoretic paradoxes from the potentialist point of view, 
making the case that the `height potentialist' perspective implied by modal structuralism allows for a more satisfying solution 
to the set-theoretic paradoxes,
 as well as clearer justifications for certain standard axioms of set theory, 
than are afforded by the more common `actualist' view. 
Papers five and six are related to nominalism, 
and (among other things) are concerned to defend the particular form of nominalism implied by his modal structuralism.

Part II offers a series of four papers 
exploring the nature and value of predicative foundations for mathematics. 
The first paper, \emph{Predicative Foundations of Arithmetic}, 
is joint with Solomon Feferman. 
It develops the first-order theory of arithmetic and proves its categoricity 
using an axiom system they call $\mathsf{EFSC}$, 
the `elementary theory of finite sets and classes',
which is argued also to be predicatively acceptable.
The framework developed presents a beatuifully economic account 
of the natural number structure from a remarkably austere collection of principles. 
In addition, the presented results may also be of relevance to contemporary discussions in philosophy.
The recent discussion of `internal categoricity' results and their philosophical significance, 
for instance, 
tends to take it as read that \emph{impredicative} comprehension is needed
to secure categoricity in this way, 
whereas the work done by Feferman and Hellman
seems to suggest that one can do with less.
A little bit -- 
in the form of $\mathsf{EFSC}$ -- 
goes further than one might expect. 

The second paper of part II, 
\emph{Challenges to Predicative Foundations of Arithmetic},
is also joint with Feferman.
It considers various philosophical objections raised to predicativism in the light 
of the mathematical work done in the first paper, 
and also presents some further technical developments contributed by Peter Aczel. 
The third paper, \emph{Predicativsm as a Philosophical Position},
is sole-authored. 
Perhaps unsurprisingly, 
it presents a more critical examination of predicativism as a philosophical position,
arguing that ultimately the central contribution of predicative foundations
is to emphasize the 
\emph{dispensability} of the uncountable from the point of view of modern science,
rather than providing an independently coherent and plausible foundation for mathematics.
The fourth paper, \emph{On the G\"odel-Friedman program}, 
is newly written for the volume.
It continues the analysis of the previous article in a more general context,
in particular that of Harvey Friedman's program for justifying large cardinal axioms 
by appeal to their utility in solving undecidable problems at the level of the naturals. 
Beyond the insightful discussion of Friedman's program, 
the article moves into novel and interesting territory regarding
the epistomology of large cardinals,
and in particular sets out the outlines of a Bayesian approach to justification here.

Part III is a little less homogeneous than the previous two parts. 
It presents five papers in the domain of philosophical logic. The first, 
\emph{Logical Truth by Linguistic Convention}, 
offers a defense of the Carnapian thesis 
that first-order logical knowledge is in a significant sense `analytic'. 
The second, \emph{Never say ``Never''!}, 
argues that intuitionism is an important sense not self-contained, 
since it turns out to rely on a `non-intuitiontistic' understanding of infintary quantification.
The third paper, 
\emph{Constructive Mathematics and Quantum Mechanics: Unbounded Operators and the Spectral Theorem} 
continues the assault on intuitionism by arguing that constructive mathematics 
is incapable of proving results needed in the foundations of quantum mechanics, 
especially the so-called `spectral theorem'. 
Paper four, \emph{If ``If-Then'' then what?}, discusses Maddy's recent attempts to revive (and extend) 
the classical position of `If-Then-ism' in foundations of mathematics.
Finally, the fifth paper, 
\emph{Mathematical Pluralism: The Case of Smooth Infinitesimal Analysis}, 
discusses the issue of how to understand the logical connectives in SIA, 
and argues that ultimately they must be understood as only as implicitly defined 
relative to a presupposed classical meta-theory, 
thus amplifying the theme in this part that intutionism and intuitionistic theories are, 
taken independently of a classical meta-translation, 
ultimately inadequate.

As this summary makes clear, the book addresses a huge 
(and mouthwatering) 
selection of topics, too many to sensibly discuss here. Instead, 
I will now turn to a discussion of two points of interest 
arising in two of the newly written papers published in the volume.

\section{On the justification of infinity}

One of the central claims of the fourth paper of part I, 
\emph{Extending the Iterative Concept of Set}, 
is that the `height potentialist' in set theory 
is able to offer a better justification for 
the axioms of infinity and replacement
than is avaiable on the standard `height actualist' framework.

The stage is set with Boolos' famous paper,
\emph{The iterative concept of set},
as the backdrop. There, 
Boolos shows how certain axioms of standard set theory 
can be motivated by appeal to 
the idea that sets are to be thought of as 
`formed in a well-ordered series of stages',
wherein at each `stage' of the process of set-formation 
one forms all possible sets 
of things formed at prior stages. 

However,
not all axioms of standard set theory were derivable 
from Boolos' axioms. Saliently, Boolos abandons the axiom of replacement 
altogether; more subtly, in order to secure infinity, 
he has to write in the existence of an infinite stage as a new axiom 
in the stage theory.

Like all standard axiomatic set theories, Boolos' stage theory is
\emph{(height) actualist}. This means that it doesn't make room for 
a substantive notion of a \emph{merely possible} set or stage of set formation.
Rather, there simply are the sets and stages of formation that there are;
there is no sensible question to ask about whether or not there might in addition be
more possible sets or stages.

The \emph{(height) potentialist} framework, on the other hand, 
says that no matter what sets there are there could always be ones of higher rank;
in terms of stages, this says that 
no matter what stages of construction have been carried out,
more such stages are possible. The idea is one of a never ending, 
`potential hierarchy' of sets, as opposed to the standard actualist 
idea of The Cumulative Hierarchy.

The paper at issue makes the striking suggestion that height potentialism
might be able to do better 
when it comes to justifying infinity and replacement 
than Boolos or the height actualist more generally is able to. 
Now as a matter of fact I agree with Hellman's claim here;
but I am not convinced by his argument.
Let me explain why. For the sake of simplicity, I will focus my discussion on infinity; 
analogous though more tedious-to-state points stand to be made 
about replacement as well.

How might height potentialism help with deriving the axiom of infinity? 
First, it is important to note that in one very mundane (but important) sense,
it necessarily cannot help. For the standard axiom of infinity says, 
simply and unapologetically, 
that there \emph{exists} and infinite set. And many height potentialist theories, 
Hellman's own brands included, 
tend not to have this as a consequence, and inded they tend not 
to even imply there exist \emph{any sets at all}. 
On potentialist theories, in general, sets are possibilia, 
and so the relevant `axiom of infinity' does not assert that 
infinite sets \emph{exist}, but rather 
only the `potentialist translation' of the standard axiom, 
that infinite sets \emph{are possible}. 
This latter is really the principle 
whose justification is at issue in the paper.\footnote{ 
    Hellman does not this in the article, but in a rather off hand way;
    I think perhaps it might have been emphaszied more, and earlier.
    }

Now is probably a good time to explain why 
\emph{I} 
believe that the height potentialist 
has available a justification for the axiom of infinity 
not available to the actualist. 
That is simply because their axiom is a 
\emph{ weakening } 
of the corresponding actualist principle. 
The height potentialist only needs to secure the 
\emph{possible} 
truth 
(in whatever sense is at issue) 
of the actualist's axiom. So obviously routes of justification 
will be open to one but not the other, though exactly how this plays out 
will depend on the brand of potentialism (and the sense of possibility) at issue.

But this simple thought is not Hellman's. 
According to Hellman, 
there are principles statable in the language of his modal stage theory 
-- essentially, that of modal, plural logic with 
vocabulary for stages and sets --
that are at once significantly more self-evident
than the potentialist's axiom of infinity 
and that nevertheless imply it as a consequence. 
It is this claim I am somewhat doubtful of.

Hellman considers one axiom of this kind 
that he ultimately rejects on grounds of its being `too close' 
to the sought conclusion to really gain any ground. This is the assertion:
\begin{description}
\item[A] It is possible that there are some stages $ss$ 
such that the empty set is formed at one of them, 
and every stage in the $ss$ is succeeded by another stage in the $ss$.\footnote{
    Following Hellman, I will use double letters $xx$ to range over pluralities, 
    and will tend to use $s$, $ss$ when the intended range is stages.
    }
\end{description}
In the context of the rest of the stage theory, 
and in particular the axiom that every stage contains all possible sets 
from among things at the previous stage,
this readily implies there is a least stage with infinitely many stages before it, 
and with it many infinite sets. But, according to Hellman, this axiom is 
\begin{quote} ``Too close for comfort to the conclusion sought. 
    Once you have an infinity of compossible things... 
    not surprisingly you generate a set of those things at the next stage.''\end{quote}
What then is Hellman's alternative? To introduce it, 
we need the following definition. 
Say that a property $P(ss)$ that applies to stages $ss$ is
 \emph{Indefinitely Extendible} (IE) 
iff each of

\begin{description} 
    \item[B]    $P(ss)$ is possible,
    \item[C]    given any possible stages $ss$ with $P(ss)$, 
                it is possible to find a $t$ after all $ss$ with $P(ss + t)$
\end{description}
are true. In the latter of course $ss + t$ is just 
the $ss$ together with $t$ (which will be supplied by 
our plural comprehension axioms). The \emph{rough} idea is that $P$ is 
an IE property when it is always possible to find further 
stages to which it applies: so no matter how far you go on 
in the process of set formation, you will always eventually 
find new series of stages that satisfy $P$.

Now, let $N$ be the predicate
\begin{quote} Every stage in the $ss$ is one at which a natural number is formed
\end{quote} 
Then the axiom Hellman proposes is simply that `$N$ is not IE'. 
Let us see what this amounts to. Substituting $N$ for $P$ in $IE$ 
and negating we get the negation of one of {\bf B} or {\bf C} with $N = P$. 
The resulting negation of {\bf B} is demonstrably false 
given the rest of Hellman's stage theory, 
so we are left with (up to trivial grammatical reshuffling)

\begin{description}
    \item[D] possibly, there are stages $ss$ 
    where every one of $ss$ witnesses the formation of a new natural number, 
    and such that no later stage witnesses the formation of a natural number.
\end{description}

Using the fact that the rest of Hellman's theory implies 
it is possible for there to be a stage $s_n$ at which 
each particular natural number $n$ is formed, 
one can easily go on to show that there must possibly be a stage 
at which the set of all natural numbers is formed (and much more besides). 
But this is the potentialist version of the axiom of infinity we set out to secure.

I have two issues with this argument. 
The first is that it is not obvious to me 
that there is much daylight between Hellman's proposed axiom 
and axiom {\bf A}. The second is that the machinery of stage theory 
and the Boolosian iterative conception seems superfluous, 
and indeed unnecessarily complicating. The indefinite extensibility property,
and formulation of axioms in terms of it, is well known in the literature 
and provides a much cleaner formulation of the same ideas, in my opinion at least.

On the issue of daylight between {\bf A} and Hellman's proposal: 
Let's review the dialectic. Hellman's stage theory is committed to 

\begin{description} 
    \item[E] For each natural number $n$, 
    it is possible to find a stage $s_n$ at which the 
    finite Von Neumann ordinal $n$ is formed.\footnote{
        This is a slightly loose and tendentious way to put things,
        since it appeals to natural numbers in the metatheory. But
        the same points can be made, in a more subtle way, without
        this crutch.
    }
\end{description} 
The question is whether in addition it is possible to form 
all such stages `at once'. {\bf A} says `yes' quite straightforwardly: 
it asserts that it is possible to run through all the finite stages 
of the set construction process. In the context of {\bf E} on the other hand, 
{\bf D} is a way of saying `yes' bentbackwardly, by saying that eventually 
the stages that witness formation of natural numbers \emph{stop coming}, 
even though the stages continue. But (given that by {\bf E} such stages 
do come for each $n$) this really seems a trivial restatement of the 
claim that eventually you get all of them.  In fact they seem roughly akin 
to the pair `someone is wearing a hat' and `not everyone is not wearing a hat': 
you can either say eventually you get all of the stages at which natural numbers 
are formed, or that eventually you won't get any more stages at which natural 
numbers are formed, but in either case you are saying essentially the same thing.

As to the second point, it is a little unclear why talk of stages 
should be relevant to such arguments for the justification of infinity, 
and in fact there are simpler though intimately related proposals known 
in the literature. For example, a simpler definition of an IE predicate says 
that the predicate $P$ is IE if necessarily, given any plurality $pp$ of $P$-ers, 
it is possible to find a $p$ not among the $pp$ with $P(p)$. 
Then the axiom that says the predicate `is a natural number' or 
`is a hereditarily finite set' is not IE in this sense seems to have 
all the plausibility of Hellman's proposal but without need to refer to stages. 
In fact, this (more or less) is the statement of the axiom of infinity in 
Linnebo's potentialist system.

Remember: note the potential that the argument for replacement does more,
because it simplifies Linnebo's axiom. This is not evident, 
cos of weirdness in defining `function'.

\section{On Bayesianism and Large Cardinals}

The second theme I wanted to pick up on in this review concerns an interesting
strand of argument in the book that plays out in part II concerning justification
of belief in the existence of large cardinals. 

One possible path to justification here, that has its roots in ideas of G\"odel, 
is that one may be able to justify belief in large cardinals by 
a kind of intra-matheamtical indispensability argument. If one can show that there
is a significant class of important `low level' mathematical problems 
(say at the level of the integers) that are unsolvable except
by appeal to large cardinals, then, the thought goes, one might have a sort of 
case to make that we should accept the existence of large cardinals
analogous to the Quine-Putnam indispensability arguments for the existence 
of mathematical objects \emph{simpliciter}.\footnote{
    Of course Hellman himself, like Putnam in fact, 
    ultimately finds such arguments dubious on grounds 
    that abstract mathematical objects are not needed anywhere
    in light of the possibility of a modal-structural translation 
    of mathematics.
} 
In its initial G\"odelian form, this idea was developed using consistency statements 
for theories of interest as the `important mathematical problems' left open by set theory
without large cardinals but solvable in set theory with them. This initial form is 
open to natural concerns about whether or not consistency statements are `mathematical'
enough to constitute `important mathematical problems'; but since G\"odel's time,
many further results have been found, including the Paris Harrington theorem 
along with a whole litany of results of Harvey Friedman, that seem more 
legitimately mathematical statements at the level of the integers, and yet
that can (in a sense!) \emph{only} be resolved by appeal to fairly strong assumptions, 
e.g. the existence of $n$-Mahlo cardinals for each natural $n$.

Now I parenthetically remarked `in a sense' above, for the reason that
this is typically only so when it comes to consistency strength. 
In a case the receives extensive discussion in CITE, for example, it is shown 
that a certain proposition in Boolean relation theory can be proven assuming 
the existence of an $n$-Mahlo cardinal for each $n$, and that this hypothesis 
is necessary in the sense that the natural large cardinal hypotheses of 
weaker strength-- saliently, for example, the scheme in $n$ to the effect 
that there exists an $n$ Mahlo cardinal for each $n$ -- cannot prove the theorem, 
on grounds that the theorem implies the $1$-consistency of such hypotheses. But 
of course this \emph{does not} mean the large cardinal hypothesis really is necessary.
In fact it is sufficient to assume that the large cardinal hypothesis in question 
only has true arithmetic consequences (essentially, that it is $1$-consistent). 
In his work, Solomon Feferman repeatedly appealed to such observations 
as a way of defusing attempts at justifying belief in uncountable infinities;
in the present context, one might see them as a way of defusing the kind of 
intra-mathematical indispensability argument under consideration.

The last two papers of part II, and especially the last one, contains the germ
of an idea for how to overcome this kind of problem by appeal to Bayesian 
confirmation theory.

The central suggestion is that the famous `Bayes rule'

\begin{equation}
P(H|E) = \frac{P(E|H)\cdot P(H)}{P(E)}
\end{equation}

may somehow be invoked to give a `boost' to the subjective probability 
of large cardinal hypotheses, even if of course that boost will fall 
short of the security provided by a deductive proof. 
This idea has roots, according to Hellman, in a famous remark of G\"odel's, 
in which G\"odel proposes that large cardinal axioms and other axioms 
extending ZFC may be justified in terms of their `success' and 
`verifiable consequences'.

Hellman's idea is to appeal to results like those of Friedman 
to `increase the posterior rational credence' of the hypothesis that large 
cardinals exist, \emph{via}  Bayes rule. According to Hellman, 
in the role of evidence (the $E$ in Bayes' rule) we may take syntactic 
claims like (say) the schema asserting the existence of an $n$ Mahlo cardinal 
for each $n$ is 1-consistent; this, says Hellman, is a proposition to which 
we have `epistemic access', `arising in a variety of contexts' in such a way that
plausibly we can hope to `conrim... it through its own fruitfulness in other applications'.
(This is in contrast e.g. to Friedman's proposition from Boolean relation theory,
which Hellman does not think has these properties.) Hellman adds that besides this
consistency claim, in the case at hand, we may throw in additionally `consistency 
properties of other cardinals whose existence is derivable from the Mahlos', 
including inaccessibles, hyper-inaccessibles and so on -- 
to further enrich the evidence. In the role of the hypothesis $(H)$ of course
we will just have the existence of an $n$-Mahlo cardinals for each $n$. 

Hellman never actually runs through how the application of Bayes rule in 
such cases is supposed to go, but I think it is illuminating to do so 
-- and reveals why application of the rule in mathematical cases in particular 
is perhaps more problematic than traditional empirical applications. 
In this case, it is clear that $P(E|H)$ is 1, so this drops out. 
We are left with the claim that one's degree of belief in the claim that 
the $n$-Mahlo cardinals exist, given they are 1-consistent, should be at least 
the result of dividing one's degree of belief that they exist by 
one's degree of belief that the corrsponding scheme is 1-consistent. (Technically
one could add to the $E$ side the other consistency properties mentioned by Hellman,
but these would be redundant for the obvious reason that they are all implied by 
the consistency property already mentioned, and so conjoining them
will not alter the probability of $E$.)

At this point we run into a bit of a problem. If one's degree of belief 
in the 1-consistency of the relevant scheme is high -- which for most people 
of course, it is -- then updating on this will make little difference to the 
posterior on their existence. Few people then will have their credences in 
Mahlos altered by this line of reasoning. Certainly, a skeptic like Feferman already
willing to concede consistency, but doubtful only of existence, will be little moved.

Now to \emph{some} extent this might be chalked up simply to the problem of 
`old evidence', familiar from discussions of Bayseianism. This is the 
problem that for Bayesian update to have \emph{any} effect, there must be a 
change in view on the $E$. Consider for example $E =$ `the die comes up 6', 
$H =$ `the die is loaded', against some background information $B$. Suppose 
the die has been rolled; you are basically certain its a six. Then applying 
Bayes rule will yield the wrong results; really you want $P(E) = 1/6$, 
not close to $1$. In this case, and with other similar straightforward cases, 
it is obvious what to say: just imagine yourself in a situation before you were 
certain in $E$, before you got the relevant evidence, and reason from there. 
But in the mathematical case, the same move looks a little less plausible. It 
is not as though we can all remember a time before we thought Mahlo cardinals 
were consistent; most people probably have taken it as read since they were 
introduced to them. It is even fairly unclear what might motivate an upward shift 
in one's credence in the consistency of Mahlos (although sharply downward is 
easy enough to conceive).

One natural idea suggested briefly by Hellman is that evidence comes from 
familiarity with the systems; ones confidence in the consistency of a system 
is gradually increased as one uses it, becomes familiar with it, all the while 
not finding any contradictions. Perhaps this is so, but in the case he discusses 
(Mahlo cardinals) it seems highly implausible that these kinds of enquires have 
significantly raised peoples confidence. Similarly inaccessibles.

Perhaps, however, there are some such cases. 
Measurables were fairly widely held to be inconsistent but gradual 
familiarity seems to have persuaded people otherwise. Similarly with the 
hierarchy of choiceless cardinals. Initially suspected to be consistent, 
but pursuit revealed rich structure theory. Increasingly people think they 
are consistent. But this consistency \emph{is} surprising. So perhaps people 
are more willing to take the case for these cardinals -- and hence the case 
for the ultimate falsity of AC -- more seriously. In this case, Hellman's 
Bayseian model does seem to apply in a potentially interesting way.

%%%%% Old work Below
\section{Scraps}
I want to raise some issues surrounding 
the fourth paper of this part, 
which is newly written for the volume. 
In seminal work CITE, George Boolos showed 
how some standard axioms of set theory could 
be recovered from a theory of sets according 
to which they are ``formed in stages''. 
The paper was highly influential in the 
development of the iterative concept of 
set and (I believe) its uptake among philosophers. 
However, Boolos' paper argues that the axiom of 
replacement is special in that it is not 
straightforwardly recoverable from the stage theory. 
Similarly, the axiom of infinity cannot be recoverable 
from the pure stage theory, 
and so the existence of an infinite stage must 
simply be demanded as an extra axiom. 

The fourth paper argues that the `height potentialist' 
framework, something implied by Hellman's modal 
structuralism, allows an explicit argument for both 
the axiom of infinity and the axiom of replacement 
from the point of view of the iterative concept of set. 
I want now to briefly discuss this argument, and raise 
some questions about it. For the sake of simplicity, I 
will focus my discussion on the axiom of infinity, though 
everything I say also applies to replacement.

Recall that a \emph{height actualist} in set theory is 
someone who thinks there are some sets (plural) such that 
there can be no set of higher rank than all of them; in 
terms of the stage theory, this amounts to the claim that 
there are some stages in the process of set formation such 
that there can be no stage after all of them. Intuitively 
speaking, the sets/stages in question represent the 
sets/stages of The Cumulative Hierarchy, the structure 
containing all possible sets. Height actualism was surely 
held, implicitly at least, by Boolos and many after him. 
A \emph{height potentialist}, on the other hand, is 
someone who thinks that there can always be sets of 
higher rank, or that there can always be further 
stages in the process of set formation. For 
the height potentialist, there is no such thing as The 
Cumulative Hierarchy, since any cumulative hierarchy 
structure can be extended to include further stages of 
set formation.

Now, how does height potentialism help with the 
derivation of the axioms of infinity? 
First, we should note that the standard axiom of infinity 
-- that there \emph{exists} an infinite set, or 
least infinite ordinal, or whatever -- is \emph{not} 
Hellman's target; rather, he aims to show that one can 
establish the \emph{possible} existence of an infinite 
set in a potentialist stage theory. Effectively, 
Hellman's claim is that there are intuitively 
plausible axioms one can adopt in the potentialist 
setting that allow one to derive the possible 
existence of an infinite set. Of course, for 
this to work, the offered axioms have to be more 
intrinsically plausible than the simple assertion 
that infinite sets / stages in the process of set 
formation are possible.

Hellman considers one axiom of this kind 
that he ultimately rejects on grounds of its being `too close' 
to the sought conclusion to really gain any ground. This is the assertion:
\begin{description}
\item[A] It is possible that there are some stages $ss$ 
such that the empty set is formed at one of them, 
and every stage in the $ss$ is succeeded by another stage in the $ss$.\footnote{
    Following Hellman, I will use double letters $xx$ to range over pluralities, 
    and will tend to use $s$, $ss$ when the intended range is stages.
    }
\end{description}
In the context of the rest of the stage theory, 
and in particular the axiom that every stage contains all possible sets 
from among things at the previous stage,
this readily implies there is a least stage with infinitely many stages before it, 
and with it many infinite sets. But, according to Hellman, this axiom is 
\begin{quote} ``Too close for comfort to the conclusion sought. 
    Once you have an infinity of compossible things... 
    not surprisingly you generate a set of those things at the next stage.''\end{quote}
What then is Hellman's alternative? To introduce it, 
we need the following definition. 
Say that a property $P(ss)$ that applies to stages $ss$ is
 \emph{Indefinitely Extendible} (IE) 
iff each of

\begin{description} 
    \item[B]    $P(ss)$ is possible,
    \item[C]    given any possible stages $ss$ with $P(ss)$, 
                it is possible to find a $t$ after all $ss$ with $P(ss + t)$
\end{description}
are true. In the latter of course $ss + t$ is just 
the $ss$ together with $t$ (which will be supplied by 
our plural comprehension axioms). The \emph{rough} idea is that $P$ is 
an IE property when it is always possible to find further 
stages to which it applies: so no matter how far you go on 
in the process of set formation, you will always eventually 
find new series of stages that satisfy $P$.

Now, let $N$ be the predicate
\begin{quote} Every stage in the $ss$ is one at which a natural number is formed
\end{quote} 
Then the axiom Hellman proposes is simply that `$N$ is not IE'. 
Let us see what this amounts to. Substituting $N$ for $P$ in $IE$ 
and negating we get the negation of one of {\bf B} or {\bf C} with $N = P$. 
The resulting negation of {\bf B} is demonstrably false 
given the rest of Hellman's stage theory, 
so we are left with (up to trivial grammatical reshuffling)

\begin{description}
    \item[D] possibly, there are stages $ss$ 
    where every one of $ss$ witnesses the formation of a new natural number, 
    and such that no later stage witnesses the formation of a natural number.
\end{description}

Using the fact that the rest of Hellman's theory implies 
it is possible for there to be a stage $s_n$ at which 
each particular natural number $n$ is formed, 
one can easily go on to show that there must possibly be a stage 
at which the set of all natural numbers is formed (and much more besides). 
But this is the potentialist version of the axiom of infinity we set out to secure.

I have two issues with this argument. 
The first is that it is not obvious to me 
that there is much daylight between Hellman's proposed axiom 
and axiom {\bf A}. The second is that the machinery of stage theory 
and the Boolosian iterative conception seems superfluous, 
and indeed unnecessarily complicating. The indefinite extensibility property,
and formulation of axioms in terms of it, is well known in the literature 
and provides a much cleaner formulation of the same ideas, in my opinion at least.

On the issue of daylight between {\bf A} and Hellman's proposal: 
Let's review the dialectic. Hellman's stage theory is committed to 

\begin{description} 
    \item[E] For each natural number $n$, 
    it is possible to find a stage $s_n$ at which the 
    finite Von Neumann ordinal $n$ is formed.\footnote{
        This is a slightly loose and tendentious way to put things,
        since it appeals to natural numbers in the metatheory. But
        the same points can be made, in a more subtle way, without
        this crutch.
    }
\end{description} 
The question is whether in addition it is possible to form 
all such stages `at once'. {\bf A} says `yes' quite straightforwardly: 
it asserts that it is possible to run through all the finite stages 
of the set construction process. In the context of {\bf E} on the other hand, 
{\bf D} is a way of saying `yes' bentbackwardly, by saying that eventually 
the stages that witness formation of natural numbers \emph{stop coming}, 
even though the stages continue. But (given that by {\bf E} such stages 
do come for each $n$) this really seems a trivial restatement of the 
claim that eventually you get all of them.  In fact they seem roughly akin 
to the pair `someone is wearing a hat' and `not everyone is not wearing a hat': 
you can either say eventually you get all of the stages at which natural numbers 
are formed, or that eventually you won't get any more stages at which natural 
numbers are formed, but in either case you are saying essentially the same thing.

As to the second point, it is a little unclear why talk of stages 
should be relevant to such arguments for the justification of infinity, 
and in fact there are simpler though intimately related proposals known 
in the literature. For example, a simpler definition of an IE predicate says 
that the predicate $P$ is IE if necessarily, given any plurality $pp$ of $P$-ers, 
it is possible to find a $p$ not among the $pp$ with $P(p)$. 
Then the axiom that says the predicate `is a natural number' or 
`is a hereditarily finite set' is not IE in this sense seems to have 
all the plausibility of Hellman's proposal but without need to refer to stages. 
In fact, this (more or less) is the statement of the axiom of infinity in 
Linnebo's potentialist system.

Remember: note the potential that the argument for replacement does more,
because it simplifies Linnebo's axiom. This is not evident, 
cos of weirdness in defining `function'.


The second part consists in papers, some of which are joint with Solomon Feferman, that explore the nature and value of predicative foundations for mathematics. The first paper, which is joint with Feferman, presents technical underpinnings for a predicativist theory based in a modest and elegant theory of finite sets. The second, also joint with Feferman, considers some philosophical challenges raised to the account of the first paper, and also presents some further technical developments contributed by Peter Aczel. The third paper begins a philosophical analysis of the predicativist stance, and the fourth, newly written for the volume, continues that analysis, and in addition fleshes out an idea aired briefly in the third paper, namely the application of Bayesian epistemology in philosophy of mathematics as a means for justifying belief in the existence of large cardinals. The part as a whole is remarkable in showcasing the collaborative and interdisciplinary research involving mathematicians and philosophers that makes philosophy of mathematics such an exciting field.

I want here to briefly discuss the final topic, the use of Bayseian confirmation theory to support large cardinal axioms.

In the background here is an interesting dialectic regarding justification for existence of large cardinals \emph{via} their use in solving `low level', generally arithmetic, problems. The oldest idea of this kind is G\"odel's idea about incompletenes. For example, ZFC does not prove its own consistency, a statement at the level of the natural numbers, but ZFC + there exists an inaccessible does. Thus we are compelled to `move up' the cumulative hierarchy via large cardinal axioms to solve independent problems even at the arithmetic level. More sophisticated examples have since been discovered, of course, including the Paris Harrington theorem, a finite form of Kruskal's theorem, and the litany of results of this kind offered by Harvey Friedman.

All these results, however, suffer from the same problem when viewed as arguments for the indispensability of large cardinals: namely, the \emph{existence} of the large cardinals in question is in general sufficient but unnecessary; for their mere \emph{conservativeness over the arithmetic fragment of the language}, the schema $Prov_T(A) \rightarrow A$ for arithmetic $A$, will on the one hand allow for the desired results, but on the other itself remains a statement about provability at the level of the natural numbers, not something committed to large cardinals.

A suggestion in the 3rd and 4th papers of the part is that Bayesian confirmation 
theory will be relevant. This is an interesting suggestion and one that is
 well worth further exploration. At the same time, in my view, 
 much further exploration is needed, since there remain significant 
 questions about the utility of the Bayesian approach here that are 
 not answered in the paper.

The central suggestion is that confirmation theory and especially the 
famous `Bayes rule'
\begin{equation}
P(H|E) = \frac{P(E|H)\cdot P(H)}{P(E)}
\end{equation}
may somehow be invoked to give a `boost' to the subjective probability 
of large cardinal hypotheses, even if of course that boost will fall 
short of the security provided by a deductive proof. 
This idea has roots, according to Hellman, in a famous remark of G\"odel's, 
in which G\"odel proposes that large cardinal axioms and other axioms 
extending ZFC may be justified in terms of their `success' and 
`verifiable consequences'.

Hellman's idea is to appeal to results like those of Friedman 
to `increase the posterior rational credence' of the hypothesis that large 
cardinals exist, \emph{via}  Bayes rule. According to Hellman, 
in the role of evidence (the $E$ in Bayes' rule) we may take syntactic 
claims like (say) the schema asserting the existence of an $n$ Mahlo cardinal 
for each $n$ is 1-consistent; this, says Hellman, is a proposition to which 
we have `epistemic access', `arising in a variety of contexts' in such a way that
plausibly we can hope to `conrim... it through its own fruitfulness in other applications'.
(This is in contrast e.g. to Friedman's proposition from Boolean relation theory,
which Hellman does not think has these properties.) Hellman adds that besides this
consistency claim, in the case at hand, we may throw in additionally `consistency 
properties of other cardinals whose existence is derivable from the Mahlos', 
including inaccessibles, hyper-inaccessibles and so on -- 
to further enrich the evidence. In the role of the hypothesis $(H)$ of course
we will just have the existence of an $n$-Mahlo cardinals for each $n$. 

Hellman never actually runs through how the application of Bayes rule in 
such cases is supposed to go, but I think it is illuminating to do so 
-- and reveals why application of the rule in mathematical cases in particular 
is perhaps more problematic than traditional empirical applications. 
In this case, it is clear that $P(E|H)$ is 1, so this drops out. 
We are left with the claim that one's degree of belief in the claim that 
the $n$-Mahlo cardinals exist, given they are 1-consistent, should be at least 
the result of dividing one's degree of belief that they exist by 
one's degree of belief that the corrsponding scheme is 1-consistent. (Technically
one could add to the $E$ side the other consistency properties mentioned by Hellman,
but these would be redundant for the obvious reason that they are all implied by 
the consistency property already mentioned, and so conjoining them
will not alter the probability of $E$.)

At this point we run into a bit of a problem. If one's degree of belief 
in the 1-consistency of the relevant scheme is high -- which for most people 
of course, it is -- then updating on this will make little difference to the 
posterior on their existence. Few people then will have their credences in 
Mahlos altered by this line of reasoning. Certainly, a skeptic like Feferman already
willing to concede consistency, but doubtful only of existence, will be little moved.

Now to \emph{some} extent this might be chalked up simply to the problem of 
`old evidence', familiar from discussions of Bayseianism. This is the 
problem that for Bayesian update to have \emph{any} effect, there must be a 
change in view on the $E$. Consider for example $E =$ `the die comes up 6', 
$H =$ `the die is loaded', against some background information $B$. Suppose 
the die has been rolled; you are basically certain its a six. Then applying 
Bayes rule will yield the wrong results; really you want $P(E) = 1/6$, 
not close to $1$. In this case, and with other similar straightforward cases, 
it is obvious what to say: just imagine yourself in a situation before you were 
certain in $E$, before you got the relevant evidence, and reason from there. 
But in the mathematical case, the same move looks a little less plausible. It 
is not as though we can all remember a time before we thought Mahlo cardinals 
were consistent; most people probably have taken it as read since they were 
introduced to them. It is even fairly unclear what might motivate an upward shift 
in one's credence in the consistency of Mahlos (although sharply downward is 
easy enough to conceive).

One natural idea suggested briefly by Hellman is that evidence comes from 
familiarity with the systems; ones confidence in the consistency of a system 
is gradually increased as one uses it, becomes familiar with it, all the while 
not finding any contradictions. Perhaps this is so, but in the case he discusses 
(Mahlo cardinals) it seems highly implausible that these kinds of enquires have 
significantly raised peoples confidence. Similarly inaccessibles.

Perhaps, however, there are some such cases. 
Measurables were fairly widely held to be inconsistent but gradual 
familiarity seems to have persuaded people otherwise. Similarly with the 
hierarchy of choiceless cardinals. Initially suspected to be consistent, 
but pursuit revealed rich structure theory. Increasingly people think they 
are consistent. But this consistency \emph{is} surprising. So perhaps people 
are more willing to take the case for these cardinals -- and hence the case 
for the ultimate falsity of AC -- more seriously. In this case, Hellman's 
Bayseian model does seem to apply in a potentially interesting way.



The third part consists in papers on philosophical logic, generally united by the fact that the issues discussed have some bearing on foundations for mathematics. Explain.

Really good book. To quote Hellman on Burgess and Rosen, philosophy at its best!

%bibliography
\bibliography{masterbib}
\bibliographystyle{asl}
%
\end{document}